---
title: "Linear regression"
subtitle: "Last assginment!"
author: "Alejandro Molina Moctezuma"
format: 
  html:
    embed-resources: true
editor: visual
---

# Instructions

You will work with three datasets. For each dataset, you must:

1.  Fit a linear model with the given response and explanatory variable(s). You need to identify which is which!
2.  Interpret the model summary, identifying $\beta_0, \beta_1$ whether each is significant, the estimate of $\sigma$ , and $R^2$ (and interpret $R^2$ in context)
3.  Use the four steps of NHST we covered in class, and see whether the explanatory variable has an effect on the response variable. Write a biological conclusion.
4.  Repeat this full set of tasks for all three datasets.
5.  All datasets (flowers, snails, wood are available on Canvas)
6.  I help you with the first one!

------------------------------------------------------------------------

# Data

## Dataset 1: More rain, more flowers? 🌻

This dataset contains observations of flower production across multiple sites that received different amounts of rainfall. Flower production is measured as a continuous count of total flowers produced at each site.

To run the model, first read the file into an object called flowers_df.

Run a linear model using:

**Goal:** Determine whether rainfall affects flower production.

### Instructions model 1

To run the model, first read the file into an object called flowers_df.

Run a linear model using:

```{r eval=FALSE}
modelflower<-lm(flower_production ~ rainfall, data = flowers_df)
```

And then, view the summary of the model using:

```{r eval=FALSE}

summary(modelflower)

```

Finally, plot the model using ggplot2:

```{r eval=FALSE}
library(ggplot2)
ggplot(flowers_df, aes(x = rainfall, y = flower_production)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Flower Production vs Rainfall",
       x = "Rainfall (mm)",
       y = "Flower Production")
```

And that is it! You ran the model, viewed the summary, and plotted the results.

------------------------------------------------------------------------

## Dataset 2: Effects of temperature on snail size 🐌

This dataset contains measurements of snail size collected across locations with different average temperatures. Snail size is a continuous measurement representing shell length.

**Goal:** Determine whether temperature affects snail size.

This one you can do it by yourself! Follow the same steps as in Dataset 1, but using the snail dataset.

------------------------------------------------------------------------

## Dataset 3: Wood Production, Temperature, and Forest Management 🌲

This dataset contains measurements of wood production collected from two types of forests: managed and unmanaged across gradient of temperature.

::: callout-caution
## Multiple explanatory variables!

Here you have both rain and management type as potential explanatory variables... Because both affect the wood production.
:::

**Goal:** Determine whether temperature and forest management type affect wood production.

### Instructions model 3

To run the model, first read the forest data file into an object called wood_df. Run a linear model using:

```{r eval=FALSE}
modelwood<-lm(wood_production ~ temperature + management, data = wood_df)
```

And then, view the summary of the model. You know how to do this.

Then plot the model, but! Add color to differentiate between management types:

```{r eval=FALSE}
library(ggplot2)
ggplot(wood_df, aes(x = temperature, y = wood_production, color =
 management)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Wood Production vs Temperature by Management Type",
       x = "Temperature (°C)",
       y = "Wood Production")
```

And that is it! You ran the model, viewed the summary, and plotted the results.

Submit it on Canvas, and get ready for the exam.
