---
title: "One sample t-test"
format:
  html:
    df-print: kable
---

💻Run this exercise in `EXCEL`

Normal human body temperature, as kids are taught in North America, is 98.6°F. But how well is this supported by data? Researchers obtained body-temperature measurements on randomly chosen healthy people (Shoemaker 1996). The data for 25 of those people are as follows:

```{r echo=FALSE}
read.csv("datatable.csv")
```

You can download the data from Canvas: `dataframe.csv`

You are testing whether the claim that 98.6 is the average body temperature. Do the following:

### Task 1

Write your null and alternative hypotheses

### Task 2

Run a t-test! You will do this in Excel. Use the following equation:

$$
t = \frac{ \bar{Y} - \mu}{S.E._\bar{y}} 
$$

You can use the `average` function to estimate the mean, the `stdev` function to estimate the standard deviation. To obtain the Standard error, you have to check the standard error equation! It is available on the last slidshow PDF I uploaded to Canvas.

Report the t value

### Task 3

Find a P-value.

#### Task 3.1 approximate P-value

Read the Statistical Tables section in the Book (page 711)

Read the Student-t distribution section (page 717)

Find an approximate P-value using the table.

Remember! We are doing a two tail test! And degrees of freedom are n-1

#### Task 3.2 Exact P-value

In Excel use:

```         
= 2*(1-T.DIST(ABS(t),df,TRUE))
```

Make sure to replace `t` and `df` with your values!

### Task 4

Make your statistical decision

### Task 5

💻Run this task in `R`

Now estimate it in R.

First, calculate the mean using `mean()`, calculate the s.d. using `sd()` and calclulate the standard error.

Second, find the real p value using:

```{r eval=F}
2*pt(q = abs(t),df = ,lower.tail = FALSE)
```

Replace t by your t value, and add degrees of freedom.

#### Task 5.1 Run the whole thing as a test in R

Do the following:

```{r eval=FALSE}
t.test(res$temperature,mu=98.6)
```

Replace x by a vector containing the temperatures.

## Submit

Submit your excel file and your R code either as an r file or you can copy and paste it in Canvas
