---
title: "One sample t-test"
subtitle: "AGNR/SNR 220 comparing means - one sample"
author: "Dr. Alejandro Molina"
format:
  html:
    df-print: kable
    embed-resources: true
---

💻 Run Tasks 1–4 in **Excel**. If you do not have a laptop, you can complete these tasks by hand with a calculator, except Task 3.2. Complete Task 3.2 and the R tasks when you have access to a computer.

Normal human body temperature, as kids are taught in North America, is 98.6°F. But how well is this supported by data? Researchers obtained body-temperature measurements on randomly chosen healthy people (Shoemaker 1996). The data for 25 of those people are as follows:

```{r}
#| echo: false
res <- read.csv("datatable.csv")
knitr::kable(res)
```

You can download the data from Canvas: `datatable.csv`

You are testing whether the population mean body temperature differs from 98.6°F. Use a significance level of $\alpha = 0.05$.

## Task 1 
#### Excel 💻 or by hand

Write your null and alternative hypotheses. Define what $\mu$ represents.
Write them in words and in symbols.

## Task 2
#### Excel 💻 or by hand


Calculate your t value! Use the following equation:

$$
t = \frac{\bar{Y} - \mu_0}{SE_{\bar{Y}}}
$$

Here, $\mu_0$ is the population mean specified by the null hypothesis.

In Excel, use `AVERAGE()` to calculate the sample mean and `STDEV.S()` to calculate the sample standard deviation. If you are working by hand, calculate these using your notes and a calculator.

Calculate the standard error using:

$$
SE_{\bar{Y}} = \frac{s}{\sqrt{n}}
$$

Here, $s$ is the sample standard deviation and $n$ is the sample size. In Excel, use `SQRT()` to calculate the square root.

Report:

- Sample size.
- Sample mean.
- Sample standard deviation.
- Standard error.
- t value.

Keep several decimal places during your calculations. Round your final t value to three decimal places.

## Task 3

Find a P-value.

### Task 3.1 Approximate P-value

**Read these instructions before opening the tables.**

The t distribution is symmetric and bell-shaped, with heavier tails than the normal distribution. These heavier tails account for the uncertainty from estimating the population standard deviation using a sample.

Its shape depends on the **degrees of freedom**, so different degrees of freedom require different tables. With more degrees of freedom, the t distribution gets closer to the standard normal distribution.

For a one-sample t-test:

$$
df = n - 1
$$

#### How to use the tables

1. Calculate your degrees of freedom and open the matching tab.
2. Find the absolute value of your t statistic, $|t|$. Use its distance from zero, regardless of whether it is positive or negative.
3. Compare $|t|$ with the critical t values in the table.
4. Read the corresponding **two-tailed probabilities**. If your value falls between two critical t values, your P-value falls between their probabilities.

For example, if $|t|$ falls between the critical values for probabilities of 0.10 and 0.05, report:

$$
0.05 < P < 0.10
$$

If $|t|$ is smaller than the first critical value, report $P > 0.50$. If it is larger than the last critical value, report $P < 0.001$.

**Remember! We are doing a two-tailed test.** The probabilities in these tables already include both tails. You do not need to double them.

Each graph shows the rejection regions for $\alpha = 0.05$. Each tail contains 0.025 of the total area. A t value in either shaded region leads us to reject the null hypothesis at this significance level.

**Report your degrees of freedom and the approximate P-value range from the appropriate table.**

*Click the box to open tha tables, after you have read the instructions*

:::: {.callout-note collapse="true"}
## Show or hide the t-tables and graphs

::: {.panel-tabset}

```{r}
#| echo: false
#| results: asis
#| fig-width: 7
#| fig-height: 4
#| out-width: "100%"

df_values <- c(5, 10, 15, 20, 22, 24, 25, 30, 35, 40, 60, 100, 200)
two_tailed_p <- c(
  0.50, 0.40, 0.30, 0.25,
  0.20, 0.19, 0.18, 0.17, 0.16,
  0.15, 0.14, 0.13, 0.12, 0.11,
  0.10, 0.09, 0.08, 0.07, 0.06,
  0.05, 0.045, 0.04, 0.035, 0.03,
  0.025, 0.02, 0.015, 0.01,
  0.008, 0.006, 0.004, 0.002, 0.001
)

for (d in df_values) {
  cat("\n\n#### df = ", d, "\n\n", sep = "")

  cat(
    "Use this table when your degrees of freedom equal ", d,
    ". Compare your absolute t value with the critical values below ",
    "and read across to find the two-tailed probability.\n\n",
    sep = ""
  )

  t_table <- data.frame(
    "Critical |t|" = sprintf("%.3f", qt(1 - two_tailed_p / 2, df = d)),
    "Two-tailed probability" = sprintf("%.3f", two_tailed_p),
    check.names = FALSE
  )

  cat(knitr::kable(t_table, format = "pipe", align = c("r", "r")),
      sep = "\n")
  cat("\n\n")

  critical_t <- qt(0.975, df = d)
  x <- seq(-5, 5, length.out = 2001)

  plot(
    x, dt(x, df = d),
    type = "n",
    main = paste("t distribution with", d, "degrees of freedom"),
    xlab = "t value",
    ylab = "Density",
    ylim = c(0, 0.45),
    bty = "l"
  )

  left_tail <- seq(-5, -critical_t, length.out = 500)
  right_tail <- seq(critical_t, 5, length.out = 500)

  polygon(
    c(left_tail, rev(left_tail)),
    c(dt(left_tail, df = d), rep(0, length(left_tail))),
    col = "#F4A582", border = NA
  )

  polygon(
    c(right_tail, rev(right_tail)),
    c(dt(right_tail, df = d), rep(0, length(right_tail))),
    col = "#F4A582", border = NA
  )

  lines(x, dt(x, df = d), lwd = 2)

  abline(
    v = c(-critical_t, critical_t),
    lty = 2, col = "#B2182B"
  )

  legend(
    "topright",
    legend = c(
      "Shaded tails: rejection regions",
      paste("Critical values: ±", round(critical_t, 3))
    ),
    fill = c("#F4A582", NA),
    border = NA,
    bty = "n",
    cex = 0.8
  )

  cat("\n\n")
}
```

:::
::::

### Task 3.2 Exact P-value
#### Excel 💻 

In Excel, use:

```text
=T.DIST.2T(ABS(t),df)
```

Replace `t` and `df` with your values or the cells containing them.

This function gives the two-tailed P-value.

If you are working by hand, complete this step later when you have access to a computer.

Report your P-value. Does it fall within the range you found using the table?

## Task 4

Make your statistical decision using $\alpha = 0.05$.

Do you reject or fail to reject the null hypothesis? Explain what your decision means for the claim that the population mean body temperature is 98.6°F.

## Task 5
#### 💻 Run this task in **R**

::: {.callout-warning}
## Download the data again before starting in R

Re-download `datatable.csv` from Canvas and save it in your R project folder. **Do not use the file you modified in Excel**, as those changes can cause issues when reading and analyzing the data in R.
:::

First, read the data into an object called `res`. Make sure `datatable.csv` is in your project folder.


```{r}
#| eval: false

library(readr)

res <- read_csv("datatable.csv")

# Check the column names and the first few rows.
names(res)
head(res)
```


::: {.callout-important}
## Calculate it step by step

For this task, calculate the mean, standard deviation, standard error, t value, and P-value step by step. Do not use `t.test()` yet! You will use it in Task 5.1 to check your calculations.
:::

Calculate the mean using `mean()`, calculate the standard deviation using `sd()`, and calculate the standard error.

Then calculate your t value.

```{r}
#| eval: false

n <- length(res$temperature)

sample_mean <- mean(res$temperature)
sample_sd <- sd(res$temperature)

# Replace NA with your calculations.
standard_error <- NA
t_value <- NA
```

Find the two-tailed P-value using:

```{r}
#| eval: false

2 * pt(q = abs(t_value), df = n - 1, lower.tail = FALSE)
```

Compare your results with your earlier calculations.

### Task 5.1 Run the whole test in R
#### 💻 Run this task in **R**.

Run:

```{r}
#| eval: false

t.test(res$temperature, mu = 98.6, alternative = "two.sided")
```

Find the t value, degrees of freedom, and P-value in the output. Do they match your calculations?

## Task 6
#### 💻 Run this task in **R**.


Plot the t distribution, rejection regions, and your observed t value.

If the population mean temperature truly is 98.6°F, imagine repeatedly taking random samples of 25 people and calculating a t value for each sample. Assuming population temperatures are normally distributed, these values follow a t distribution with 24 degrees of freedom.

The shaded regions below represent the rejection regions for a two-tailed test with $\alpha = 0.05$. Each tail contains 0.025 of the total area.

Replace the first `NA` with your degrees of freedom and the second `NA` with your calculated t value.

```{r}
#| eval: false

# Replace these two values.
df <- NA
t_value <- NA

# Find the critical t value for a two-tailed test.
alpha <- 0.05
critical_t <- qt(1 - alpha / 2, df = df)

# Set the horizontal range.
x_limit <- max(4, abs(t_value) + 1, critical_t + 1)
x <- seq(-x_limit, x_limit, length.out = 2001)

# Draw the t distribution.
plot(
  x, dt(x, df = df),
  type = "n",
  main = "t distribution if the population mean is 98.6°F",
  xlab = "t value",
  ylab = "Density",
  bty = "l"
)

# Shade the rejection regions for alpha = 0.05.
left_tail <- seq(-x_limit, -critical_t, length.out = 500)
right_tail <- seq(critical_t, x_limit, length.out = 500)

polygon(
  c(left_tail, rev(left_tail)),
  c(dt(left_tail, df = df), rep(0, length(left_tail))),
  col = "#92C5DE",
  border = NA
)

polygon(
  c(right_tail, rev(right_tail)),
  c(dt(right_tail, df = df), rep(0, length(right_tail))),
  col = "#92C5DE",
  border = NA
)

# Add the distribution curve.
lines(x, dt(x, df = df), lwd = 2)

# Mark the boundaries of the rejection regions.
abline(
  v = c(-critical_t, critical_t),
  col = "gray40",
  lty = 3
)

# Mark the observed t value and its opposite.
abline(v = t_value, col = "#B2182B", lwd = 2)
abline(v = -t_value, col = "#B2182B", lty = 2, lwd = 2)

legend(
  "topright",
  legend = c(
    "Observed t value",
    "Equally extreme value",
    "Critical values"
  ),
  col = c("#B2182B", "#B2182B", "gray40"),
  lty = c(1, 2, 3),
  lwd = c(2, 2, 1),
  bty = "n",
  cex = 0.8
)

mtext(
  paste0(
    "Shaded rejection regions: alpha = 0.05; critical values = ±",
    round(critical_t, 3)
  ),
  side = 3,
  line = 0.3,
  cex = 0.8
)
```

### How to interpret your graph

The solid red line marks your observed t value. The dashed red line marks an equally extreme value on the opposite side of zero. The dotted gray lines mark the critical values.

If your observed t value falls in either shaded region, it is among the most extreme 5% of t values expected under the null hypothesis. **Reject the null hypothesis** at $\alpha = 0.05$.

If your observed t value falls between the critical values, **fail to reject the null hypothesis**. Your sample does not provide sufficient evidence that the population mean differs from 98.6°F. This does not establish that the population mean equals 98.6°F.

The shaded area is fixed at 0.05. Your P-value is the area beyond your observed t value and the equally extreme value in the opposite tail.

**Use your graph to explain your decision in two or three sentences.** Describe where your observed t value falls relative to the rejection regions, explain why you rejected or failed to reject the null hypothesis, and state what this means for the population mean body temperature.