---
categories:
  - descriptive-statistics
  - central-tendency
  - spread
  - mean-median-mode
---

# Week 2, Part 1: Describing Data in R {#sec-week02}

## What you will do

In class, you will:

1.  Create an RStudio Project.
2.  Create a Quarto document that renders to a Word document.
3.  Practice a few basic R commands.
4.  Import the systolic blood pressure dataset used in our previous class.
5.  Use R to calculate the mean, median, first quartile, third quartile, standard deviation, variance, and coefficient of variation.
6.  Create a histogram.
7.  Submit

After class, you will:

1.  Read the rest of this chapter.
2.  Complete only the sections labeled **Assignment question**.
3.  Write your R code and answers in your Quarto document.
4.  Render the document to Word and submit the `.docx` file.

::: callout-important
## What should you submit?

Today, you should submit one rendered Word document that includes your name, and answers.
:::

## Part 1: Set up your work

### Create an RStudio Project

Projects keep your data, code, and output together. They also make it easier for R to find your files.

::: callout-important
## Where to put the project

You will be downloading ALL DATA FILES to this folder. Please make it an easy to find folder
:::

1.  Open RStudio.
2.  Select **File \> New Project \> New Directory \> New Project**.
3.  Name the project `SNR220` --\> You will use this for the rest of the semester
4.  Choose a location on your computer that you can find easily.
5.  Create the project.
6.  Save the systolic blood pressure `.csv` file inside the project folder.

### Create a Quarto document

1.  Select **File \> New File \> Quarto Document**.
2.  Enter a title such as `Week 2: Describing Data`.
3.  Enter your name as the author.
4.  Select **Word** as the output format.
5.  Select **Create**.
6.  Save the file as `week2_describing_data.qmd` inside your project folder.

The top of your file should look similar to this:

``` yaml
---
title: "Week 2: Describing Data"
author: "Your name"
format: docx
---
```

### Add code and written answers

You will complete all work in the `.qmd` file. Add regular text for your written answers. Add executable R code chunks for your code.

To insert a code chunk, select **Insert \> Executable Cell \> R**. You can also type three backticks followed by `{r}`, add your code, and close the chunk with three more backticks.

For example:

```` markdown
### My first calculation

The result means that...

```{r}
2 + 2
```
````

To run one line or a selected section of code, press **Ctrl + Enter** on Windows or **Command + Enter** on macOS. To create the Word document, click **Render**.

## Part 2: Play with R

R can be used as a calculator. In your Quarto document, create an R code chunk and run each of the following commands one at a time.

``` r
5 + 5

8^2

sqrt(9)

10 / 4
```

R can also store values in objects. The assignment operator `<-` places a value inside an object.

``` r
x <- 17

y <- x * 2

x

y
```

A vector is a collection of values. The function `c()` combines values into a vector.

``` r
practice_values <- c(5, 7, 9, 15, 50)

practice_values

mean(practice_values)

sd(practice_values)
```

Change one of the values and run the code again. What happens to the mean and standard deviation?

::: callout-tip
## A note about the Console

The Console is useful for quick experiments, but your assignment code should be written in the Quarto file. Code saved in your Quarto file can be checked, corrected, rerun, and included in your final Word document.
:::

## Part 3: Guided activity with the systolic blood pressure data

You previously estimated descriptive statistics for the systolic blood pressure dataset. Now you will use R to calculate the same values.

### Import the data

Download the dataforR.csv file from Canvas

Create an R code chunk in your Quarto document and enter:

```{r eval=F}
blood <- read.csv("systolicBloodPressure.csv")
```

```{r echo=F}
blood <- read.csv("../systolicBloodPressure.csv")
```

The assignment operator `<-` stores the imported dataset in an object named `blood`.

Use these functions to inspect the dataset:

``` r
head(blood)

names(blood)

str(blood)

summary(blood)
```

- `head()` displays the first few observations.
- `names()` displays the variable names.
- `str()` shows the structure and variable types.
- `summary()` provides a quick summary of each variable.



### Calculate the descriptive statistics

Use the following functions to estimate the function in your Quarto document:

``` r
mean(blood)

median(blood)

quantile(blood, probs = 0.25)

quantile(blood, probs = 0.75)

sd(blood)

var(blood)
```

$$
CV = \frac{s}{\bar{y}} \times 100
$$

Calculate it in R:

``` r
sd(blood / mean(blood) * 100
```

::: callout-important
## Before you render and submit

Make sure your values are the same as you got on Excel
:::

Now, answer in the Quarto document:

Was it easy or hard to do this?

### Create a quick histogram

A histogram displays the distribution of a numerical variable. It can help you see the center, spread, skewness, gaps, and possible extreme values.

Add this code to your Quarto document:

``` r
hist(
  blood,
  main = "Distribution of systolic blood pressure",
  xlab = "Systolic blood pressure (mm Hg)",
  col = "lightblue",
  border = "white"
)
```

The histogram will appear automatically in the rendered Word document. Below the plot, write two or three sentences describing what you see. Consider the center, spread, shape, and any possible extreme values.

::: callout-warning
That is it! You get to go home. The assignment will be due Wednesday 2nd! After we are back from labor day!
:::

# Part 4: Concepts to review

### Center and spread

Raw observations can be difficult to interpret. Descriptive statistics reduce a dataset to a few useful summaries.

We usually want to know:

1.  Where is the center of the distribution?
2.  How spread out are the observations?

Both are important. Two groups can have the same mean but very different amounts of variation.

### Measures of center

**Mean:** Add all observations and divide by the number of observations. The mean uses every value, so extreme values can strongly affect it.

**Median:** Sort the observations and find the middle value. The median is less affected by extreme values and is often useful for skewed data.

**Mode:** The value that occurs most often. It is especially useful for categorical variables and discrete counts. A dataset can have more than one mode or no unique mode.

::: callout-tip
## Think about it

A soil scientist measures nitrate concentration in eight stream samples. One sample has a much higher concentration than the others. How might that observation affect the mean? How might it affect the median? Which value would better describe a typical sample?
:::

### Quartiles

After sorting the data:

- The first quartile, Q1, is the 25th percentile.
- The median, Q2, is the 50th percentile.
- The third quartile, Q3, is the 75th percentile.

The middle 50 percent of the observations fall between Q1 and Q3.

### Variance and standard deviation

A deviation is the distance between an observation and the mean:

$$
\text{deviation}_i = y_i - \bar{y}
$$

Some deviations are positive and some are negative. If we simply add them, they cancel and sum to zero. Variance avoids this problem by squaring the deviations.

For a sample, the variance is:

$$
s^2 = \frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n - 1}
$$

The standard deviation is the square root of the variance:

$$
s = \sqrt{\frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n - 1}}
$$

Variance is expressed in squared units. Standard deviation is expressed in the original units, so it is usually easier to interpret. A larger standard deviation indicates greater variation around the mean.

::: callout-important
## Why use n - 1?

When we use a sample to estimate variation in a larger population, sample variance divides by $n - 1$. This correction improves the estimate of population variance. For now, remember that `var()` and `sd()` in R use $n - 1$ automatically.
:::

### Coefficient of variation

The coefficient of variation expresses the standard deviation relative to the mean. Because it is a percentage, it can help compare variation among measurements with different means.

For example, a CV of 12 percent means that the standard deviation is 12 percent of the mean.

Use the CV only when a value of zero is meaningful and the mean is positive. It is not appropriate for every type of measurement.

### A short example

A rancher records daily weight gain for 10 steers:

`1.2, 1.5, 1.3, 1.8, 1.1, 1.4, 1.6, 1.2, 1.7, 1.4`

The main summaries are:

| Statistic          |              Result |
|--------------------|--------------------:|
| Mean               |         1.42 kg/day |
| Median             |         1.40 kg/day |
| Q1                 |        1.225 kg/day |
| Q3                 |        1.575 kg/day |
| Variance           | 0.0529 (kg/day)$^2$ |
| Standard deviation |        0.230 kg/day |
| CV                 |               16.2% |

The mean and median are similar, which suggests that the center is not being strongly affected by an extreme value. The standard deviation describes the typical spread around the mean. It does not guarantee that every observation will fall within one standard deviation of the mean.

## Excel reference only

::: callout-note
## You do not have to complete this section

This section is included as a reference. Do not submit Excel calculations for this assignment. Complete the assignment questions using R.
:::

If values are stored in cells `A1:A10`, Excel can calculate the following:

| Statistic                 | Excel formula                          |
|---------------------------|----------------------------------------|
| Mean                      | `=AVERAGE(A1:A10)`                     |
| Median                    | `=MEDIAN(A1:A10)`                      |
| First quartile            | `=QUARTILE.INC(A1:A10,1)`              |
| Third quartile            | `=QUARTILE.INC(A1:A10,3)`              |
| Sample variance           | `=VAR.S(A1:A10)`                       |
| Sample standard deviation | `=STDEV.S(A1:A10)`                     |
| Coefficient of variation  | `=STDEV.S(A1:A10)/AVERAGE(A1:A10)*100` |

Use `VAR.S` and `STDEV.S` when the observations are a sample. The functions `VAR.P` and `STDEV.P` are used when the data include the entire population of interest.

## Homework: R exercises

Complete both assignment questions in your Quarto document. Include the code, output, plots, and written responses. Do not calculate these answers in Excel.

### Assignment question 1: Deer observations

A wildlife biologist records the number of white-tailed deer observed during eight transect surveys:

`4, 7, 3, 9, 5, 6, 4, 10`

1.  Create a vector named `deer` containing these observations.
2.  Write R code to calculate the mean, median, Q1, Q3, range, variance, standard deviation, and CV.
3.  Create a histogram with an informative title and axis label.
4.  In two or three sentences, describe the center and spread of the observations. Mention whether the mean and median are similar.

::: callout-tip
## Remember

The R function `range()` returns the minimum and maximum. To calculate the difference between them, use `diff(range(deer))`.
:::

### Assignment question 2: Wheat yield and an extreme value

A farmer records wheat yield, in bushels per acre, for six fields:

`52, 48, 55, 61, 47, 50`

A seventh irrigated field yields 110 bushels per acre.

1.  Create a vector named `wheat` containing the original six observations.
2.  Create a second vector named `wheat_with_extreme` that includes all seven observations.
3.  For each vector, write R code to calculate the mean, median, and standard deviation.
4.  Create one histogram for each vector. Give each histogram a clear title and use the same x-axis limits so the plots can be compared fairly. You can add `xlim = c(40, 115)` inside each `hist()` function.
5.  In three or four sentences, explain how the value of 110 affected the mean, median, standard deviation, and shape of the distribution.

## Before submitting

Check that your Word document includes:

- Your name and assignment title
- All R code used to answer the assignment questions
- The numerical output from R
- All required histograms
- Written interpretations in complete sentences

Click **Render** and open the `.docx` file to make sure the code, output, and plots are visible before submitting it.